---
title: USING OPEN SOURCE DATA TO VISUALIZE BIOLOGICAL, CHEMICAL, AND PHYSICAL OCEANOGRAPHIC
  PROCESSES
author: "Robert Dellinger"
date: "5/5/2022"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: cerulean
---

```{r setup, include=TRUE, message = FALSE}
knitr::opts_chunk$set(warning=FALSE, fig.algin="center")

devtools::install_github("ropensci/rerddap")
library(oce)
library(ocedata)
library(ncdf4)
library(tidyverse)
library(lubridate)
library(tidync)
library(dplyr)
library(marmap)
library(ggplot2)
library(ggmap)
library(gganimate)
library(stringr)
library(plotly)
library(sf)
library(spData)
library(rerddap)
library(raster)
library(mapdata)

```


# Mapping with the OCE package

# Read in Datasets

The OCE package is a package written by Kelley and Richards (2018) and is focused on oceanographic mapping and plotting that meets publication quality standards. Here, we use the mapPlot() function to create maps using coastline data and visualize oceanographic processes. 

# Downloading Data

First we will read in data from the oce package using the data() function 

```{r, Downloading Data}

Coastline <- data("coastlineWorld") #coastline data
Coastline_Medium <- data("coastlineWorldMedium") #medium-scale coastline data
Coastline_Fine <- data("coastlineWorldFine") #fine-scale-scale coastline data
Bathymetry <- as.topo(getNOAA.bathy(-180, 0, 0, 90)) #coordinates for bathymetry data from NOAA (Northern Hemisphere)

```

# Creating a Map

Using the map from the oce package we will create a simple map function that can be rerun

```{r, Creating a Map Function Using OCE, fig.cap="Map of California"}

# setting the map limits
par(mar=c(2, 2, 1, 1)) # margins
lonlim = c(-126,-116) # longitude
latlim = c(32,40) # latitude

coastline.map <- function() { 
  mapPlot(coastlineWorldFine, #using fine-scale data
        projection="+proj=mill", #setting the projection type
        col="grey", #setting colors,
        longitudelim=lonlim, 
        latitudelim=latlim,
        lonlabels=TRUE,
        latlabels=TRUE,
        geographical=4,
        clip=TRUE)
}

coastline.map()
```


# Creating a Map with Bathymetry

We will now create a map that mimics the depth of the ocean floor using our previous map function. The provided color palettes in the package can be used to visualize oceanographic phenomena such as bathymetry. 


```{r, fig.cap="Map of California with Bathymetry", warning=FALSE}

# setting the map limits
par(mar=c(2, 2, 1, 1)) # margins
lonlim = c(-125,-116) # longitude
latlim = c(33.5,39.5) # latitude

#Creating a Map with Bathymetry
#Place Map
coastline.map()
#Overlaying Bathymetry Data
mapImage(Bathymetry, col=oceColorsGebco, breaks=seq(-6000, 0, 100))
mapImage(Bathymetry, col=oceColorsGebco, breaks=seq(-6000, 0, 500), filledContour = TRUE)
#Adding in Coastline
mapPolygon(coastlineWorldFine, col = "darkolivegreen3", border = "black") 
#Adding in Grid
mapGrid(dlongitude=1, dlatitude=1, col="black", lty=3) 

#Setting points on a map with names and the respective locations
sites = data.frame(name = c("Leo Carrillo State Beach \n Malibu, CA", "Bodega Marine Lab \n Bodega Bay, CA"), 
                      lon = c(-118.9279, -123.0481), #locations and their respective longitude and latitudes
                      lat = c(34.0622, 38.3332)) 
#Placing points on the map
mapPoints(longitude = sites$lon, #placing points on map using longitude
          latitude = sites$lat, #placing points on map using latitude
          pch = 20, 
          cex = 1.75)  #point size
#Overlaying text
mapText(longitude = sites$lon+1.5, #map text location long and lat
        latitude = sites$lat+.75, 
        labels = sites$name,
        cex = 1.2, #font size
        col="black") # changing font family

```


```{r, fig.cap="Map of CLeo Carillo Site with Bathymetry", warning=FALSE}

# setting the map limits
par(mar=c(2, 2, 1, 1)) # margins
lonlim = c(-125,-116) # longitude
latlim = c(33.5,34.5) # latitude

#Creating a Map with Bathymetry
#Place Map
coastline.map()
#Overlaying Bathymetry Data
mapImage(Bathymetry, col=oceColorsGebco, breaks=seq(-6000, 0, 100))
mapImage(Bathymetry, col=oceColorsGebco, breaks=seq(-6000, 0, 500), filledContour = TRUE)
#Adding in Coastline
mapPolygon(coastlineWorldFine, col = "darkolivegreen3", border = "black") 
#Adding in Grid
mapGrid(dlongitude=1, dlatitude=1, col="black", lty=3) 

#Setting points on a map with names and the respective locations
sites = data.frame(name = "Leo Carrillo State Beach \n Malibu, CA", 
                      lon = (-118.9279), #locations and their respective longitude and latitudes
                      lat = (34.0622)) 
mapPoints(longitude = sites$lon, #placing points on map using longitude
          latitude = sites$lat, #placing points on map using latitude
          pch = 20, 
          cex = 1.75)  #point size
#Overlaying text
mapText(longitude = sites$lon+.5, #map text location long and lat
        latitude = sites$lat+.5, 
        labels = sites$name,
        cex = 1.2, #font size
        col="black") # changing font family

```


```

# Creating a Map with Overlayed SST Data

We will now create a map that mimics that shows us the sea surface temperature of the ocean and use the provided temeprature color palette to visualize differences in ocean temperature. NOAA and NASA provide open source data for marine biology and oceanography data that can be used for creating visuals. For this portion we will use data downloaded from the Group for High Resolution Sea Surface Temperature (GHRSST) which has entire data sets of regional and global data linked here [https://www.ghrsst.org/]. The dataset is provided in a netcdf format so we will have to first read in the data using the nc_open function from the netcdf package (Pierce 2017) prior to graphing. 

We will download sea surface temperature (SST) analysis data provided by NOAA which provides a combination of daily satellite and in situ temperature temperature measurements on a global level. We will use data from August 1, 2018, the day of the hottest observed SST by Scripps Institution of Oceanography since the previous record set in 1931. 

# Reading in Open Source SST Data from NOAA and NASA 

```{r, Downloading Data (NOAA/NASA)}

# Opening downloaded data from NOAA 
sst.data <- nc_open("/Users/robertdellinger/Documents/Research/Data/Temperature_Data/sst.08.01.18.nc")

# extract the longitude variable
sst.lon = ncvar_get(sst.data, "lon")
# extract the latitude variable
sst.lat = ncvar_get(sst.data, "lat")
# extract sst variable
sst = ncvar_get(sst.data, "analysed_sst")
# extract time 

# Kelvin to Degree celsius and calibrate
sst = sst-273.149993896484

```


# Hottest Observed SST for the Southern California Region

```{r, fig.cap="SST Map for Souththern California", warning=FALSE}

# setting the map limits
par(mar=c(2, 2, 1, 1)) # margins
lonlim = c(-122.25,-116.25) # longitude
latlim = c(32.25,35.25) # latitude
templim = c(10,26) #temperature limits in celsisus

# draw palette (must be done before plotting)
drawPalette(zlim = templim, 
            zlab = "Temperature (°C)",
            col=oce.colorsJet, #using the oce package temperature colors
            at = seq(10,26,1), 
            pos = 1)

#adding in sst to overlay the map function
coastline.map()
mapImage(longitude = sst.lon, latitude = sst.lat, z = sst, zlim = templim, col = oceColorsJet(120))
mapImage(longitude = sst.lon, latitude = sst.lat, z = sst, zlim = templim, col = oceColorsJet(120), filledContour = TRUE)  #overlay sst & contour
mapContour(longitude = sst.lon, latitude = sst.lat, z = sst, levels = seq(10,26,1),
            col = "black", lty = 1, lwd = 1)
mapPolygon(coastlineWorldFine, col="darkolivegreen3") # overlay green topography
mapGrid(dlongitude=1, dlatitude=1, col="black", lty=3) # adding a map grid

#Setting points on a map with names and the respective locations
site = data.frame(name = c("Leo Carrillo State Beach \n Malibu, CA"), 
                      lon = -118.9279, #locations and their respective longitude and latitudes
                      lat = 34.0622)
#Placing points on the map
mapPoints(longitude = site$lon, #placing points on map using longitude
          latitude = site$lat, #placing points on map using latitude
          pch = 20, 
          cex = 1.75)  #point size
#Overlaying text
mapText(longitude = site$lon+1.5, #map text location long and lat
        latitude = site$lat+.75, 
        labels = site$name,
        cex = 1.2, #font size
        col="black") # changing font family


```

# Hottest Observed SST for the Northern California Region

```{r, fig.cap="SST Map for Northern California", warning=FALSE}

# setting the map limits
par(mar=c(2, 2, 1, 1)) # margins
lonlim = c(-126.25,-120.25) # longitude
latlim = c(37.25,40.25) # latitude 
templim = c(10,26) #temperature limits in celsisus

# draw palette (must be done before plotting)
drawPalette(zlim = templim, 
            zlab = "Temperature (°C)",
            col=oce.colorsJet, #using the oce package temperature colors
            at = seq(10,26,1), 
            pos = 1)

#adding in sst to overlay the map function
coastline.map()
mapImage(longitude = sst.lon, latitude = sst.lat, z = sst, zlim = templim, col = oceColorsJet(120))
mapImage(longitude = sst.lon, latitude = sst.lat, z = sst, zlim = templim, col = oceColorsJet(120), filledContour = TRUE)  #overlay sst & contour
mapContour(longitude = sst.lon, latitude = sst.lat, z = sst, levels = seq(10,26,1),
            col = "black", lty = 1, lwd = 1)
mapPolygon(coastlineWorldFine, col="darkolivegreen3") # overlay green topography
mapGrid(dlongitude=1, dlatitude=1, col="black", lty=3) # adding a map grid


#Setting points on a map with names and the respective locations
site = data.frame(name = c("Bodega Marine Lab \n Bodega Bay, CA"), 
                      lon = -123.0481, #locations and their respective longitude and latitudes
                      lat = 38.3332) 
#Placing points on the map
mapPoints(longitude = site$lon, #placing points on map using longitude
          latitude = site$lat, #placing points on map using latitude
          pch = 20, 
          cex = 1.75)  #point size
#Overlaying text
mapText(longitude = site$lon+1.5, #map text location long and lat
        latitude = site$lat+.75, 
        labels = site$name,
        cex = 1.2, #font size
        col="black") # changing font family


```



# Plotting Using NOAA ERDAPP Data 

For this portion we will use NOAA servers to download monthly mean temperatures and visualize data using a boxplot. Data can either be downloaded by using the rerddap package or the NOAA online repository. We will clean the data by selecting for the year 2021 and then the longitudes and latitudes of the Southern California Region. Then, we will average SST per month and create an animated plot to visualize changes between months.

[Data Linked Here](https://psl.noaa.gov/repository/entry/show?entryid=cac1c2a6-a864-4409-bb77-1fdead8eeb6e)

Available ERDDAP data can also be pulled into R using the servers() function 

# Data Wrangling

```{r, Data Wrangling SST, warning=FALSE}

sst.file <- tidync("/Users/robertdellinger/Documents/Research/Data/Temperature_Data/sst.mnmean.nc") 
sst.tibble <- hyper_tibble(sst.file,  na.rm = FALSE) #turnning nc into tibble

sst.data <- sst.tibble  %>%  
  mutate(date=as.Date(time, origin="1800-1-1 00:00:00"),
         day = day(date),  # save dates as intergers 
         month = month(date) %>% as.integer(),
         year = year(date) %>% as.integer(),
         sst=as.numeric(sst)) %>%  #converting dates 
  filter(date >= dmy(010121), date <= dmy(010122),
         lat > 30, lat < 35, #lat limits
         lon > 290, lon < 310,
         year ==2021) %>% #lon limits
  dplyr::select(date, month, lon, lat, sst)
  
sst.data = as.data.frame(sst.data) #convert to data frame

# group sst based on longitude, latitude and month
sst.month = sst.data %>% 
  group_by(lon,lat,month) %>% # selected for lat, long, and month
  summarise(sst = mean(sst, na.rm = TRUE))

# ungroup grouped data frame
sst.month = sst.data %>% ungroup()

```

# Creating an Animated Temperature Plot for the Southern California Region

Here we use gganimate and ggplot2 to create a animated boxplot using mean temperature data for the Southern California Region. 


```{r, fig.cap="Monthly Variation of Temperature in Southern California Boxplot"}

animatedplot <- ggplot(data =sst.month, #ggplot sst monthly data
       aes(x = month, y = sst, fill = as.factor(month)))+
  geom_boxplot(show.legend = FALSE)+ #create boxplot
  scale_fill_brewer(palette = "Paired")+
  cowplot::theme_cowplot()+ 
  scale_x_continuous(breaks = 1:12)+
  ylab("Sea Surface Temperature (°C)")+  #add labels
  xlab("Month") +
  labs(title = "{closest_state}")+
  transition_states(month, transition_length = 1, #transition time and month
                    state_length = 0.5, wrap = TRUE) + #transition wrap
  ease_aes('linear') +
  shadow_mark()

animatedplot

```


# References 

Grolemund, G., & Wickham, H. (2011). Dates and times made easy with lubridate. Journal of Statistical Software, 40(3), 1–25. Retrieved from http://www.jstatsoft.org/v40/i03/

Kelley, Dan. 2015. Ocedata: Oceanographic Datasets for Oce. https://CRAN.R-project.org/package=ocedata.

Kelley, Dan, and Clark Richards. 2018. Oce: Analysis of Oceanographic Data. https://CRAN.R-project.org/package=oce.

R Core Team. 2018. R: A Language and Environment for Statistical Computing. Vienna, Austria: R Foundation for Statistical Computing. https://www.R-project.org/.

Richard W. Reynolds, Viva F. Banzon, and NOAA CDR Program (2008): NOAA Optimum Interpolation 1/4 Degree Daily Sea Surface Temperature (OISST) Analysis, Version 2. [indicate subset used]. NOAA National Centers for Environmental Information. http://doi.org/doi:10.7289/V5SQ8XB5 

Pedersen, T. L., & Robinson, D. (2017). Gganimate: A grammar of animated graphics. Retrieved from http://github.com/thomasp85/gganimate

Wickham, H. (2016). Ggplot2: Elegant graphics for data analysis. Springer-Verlag New York. Retrieved from http://ggplot2.org

Pierce, David. 2017. Ncdf4: Interface to Unidata netCDF (Version 4 or Earlier) Format Data Files. https://CRAN.R-project.org/package=ncdf4.
