---
title: "Downloading and Subsetting NOAA Data"
author: "Robert Dellinger"
date: "5/16/2022"
output: html_document
---

# Install Packages 

```{r setup, include=TRUE, message = FALSE}
knitr::opts_chunk$set(warning=FALSE, fig.algin="center")

devtools::install_github("rmendels/xtractomatic")
library(lubridate)
library(tidyverse)
library(xtractomatic)
library(rerddap)
library(sp)
library(spData)
library(oce)
library(ocedata)
```

# Call NOAA servers and Download Daily Data Set

```{r}

servers() #call on server

# use result  = searchData("varname:ssh") to get data from server
daily.dataset = searchData("datasetname:1day")

daily.dataset %>% glimpse() #visualize data

```

Once you have identified the datasetname of the daily product, Use the xtracto_3D() function from xtractomatic package to download the file by specifying the geogrpahical extent and time bound of interest.


# Extracting U Values for Geostrophic Currents (erdTAugeo1day)

```{r}
lon.lim = c(-126,-116) # longitude
lat.lim = c(32,40) # latitude
time.lim = c("2012-08-01", "2010-08-12")

u = xtracto_3D(dtype = "erdTAugeo1day", xpos = lon.lim, ypos = lat.lim, tpos = time.lim)

lon = u$longitude
lat = u$latitude
time = u$time %>%as.Date()
u.data = u$data 

## obtain dimensions
lon %>% length(); lat %>% length(); time %>% length(); u.data %>% dim()

dimension = data.frame(lon, u.data[,,1]) %>% dim()

## convert the array into data frame
u.tb = data.frame(lon, u.data[,,1]) %>% 
  as_tibble() %>% 
  gather(key = "lati", value = "u", 2:dimension[2]) %>% 
  mutate(lat = rep(lat, each = dimension[1]), time = time[1]) %>% 
  select(lon,lat, time, u)

```

# Extracting V Values for Geostrophic Currents (erdTAugeo1day)

```{r}
v = xtracto_3D(dtype = "erdTAvgeo1day", xpos = lon.lim, ypos = lat.lim, tpos = time.lim)

lon = v$longitude
lat = v$latitude
time = v$time%>%as.Date()
v.data = v$data 

v.tb = data.frame(lon, v.data[,,1]) %>% 
  as_tibble() %>% 
  gather(key = "lati", value = "v", 2:dimension[2]) %>% 
  mutate(lat = rep(lat, each = dimension[1]), time = time[1]) %>% 
  select(lon,lat, time, v)

```

# Extracting Sea Surface Height Values for Geostrophic Currents (erdTAugeo1day)

```{r}
ssh = xtracto_3D(dtype = "erdTAssh1day",xpos = lon.lim, ypos = lat.lim, tpos = time.lim)

lon = ssh$longitude
lat = ssh$latitude
time = ssh$time%>%as.Date()
ssh.data = ssh$data 

ssh.tb = data.frame(lon, ssh.data[,,1]) %>% 
  as_tibble() %>% 
  gather(key = "lati", value = "ssh", 2:dimension[2]) %>% 
  mutate(lat = rep(lat, each = dimension[1]), time = time[1]) %>% 
  select(lon,lat, time, ssh)
```

#Data Wrangling and Interp

```{r}
ssh.in = oce::interpBarnes(x = ssh.tb$lon, y = ssh.tb$lat, z = ssh.tb$ssh)
dimension = data.frame(lon = ssh.in$xg, ssh.in$zg) %>% dim()

ssh.in = data.frame(lon = ssh.in$xg, ssh.in$zg) %>% 
  as_tibble() %>% 
  gather(key = "lati", value = "ssh", 2:dimension[2]) %>% 
  mutate(lat = rep(ssh.in$yg, each = dimension[1]), time = time[1]) %>% 
  select(lon,lat, time, ssh)

#stitch together the dataframe
aviso = ssh.tb %>% 
  bind_cols(u.tb %>%select(u),
            v.tb %>% select(v))

#calculate velocity 
geostrophic = aviso %>% 
  mutate(vel =sqrt(u^2+v^2))


```




#Vsiualize NOAA data (Geostrophic Flow)

```{r, fig.cap="Figure 2: geostrophic current overalid on the sea surface height"}

lonlim = c(-126,-116) # longitude
latlim = c(32,40) # latitude

ggplot()+
  metR::geom_contour_fill(data = ssh.in, aes(x = lon, y = lat, z = ssh), bins = 120)+
  metR::geom_vector(data = aviso, aes(x = lon, y = lat, dx = u, dy = v),
                    arrow.angle = 25, arrow.length = .4, arrow.type = "open")+
  metR::scale_mag(max = .75, name = "Speed", labels = ".75 m/s")+
    geom_sf(data = spData::world, fill = "grey60", col = "grey20")+
  coord_sf(xlim = lonlim, ylim = latlim)+
  # theme(legend.position = "none")+
  labs(x = NULL, y = NULL)+
  scale_fill_gradientn(name = "ssh (m)",colours = oce::oceColors9A(120), na.value = "white")+
  scale_x_continuous(breaks = seq(-126,-116, length.out = 5) %>%round(1))+
  scale_y_continuous(breaks = seq(32,40, 2))+
  guides(fill = guide_colorbar(title = "Sea surface height (m)", 
                               title.position = "right", title.theme = element_text(angle = 90), 
                               barwidth = 1.25, barheight = 10, draw.ulim = 1.2, draw.llim = 0.6))+
  theme_bw()+
  theme(axis.text = element_text(colour = 1, size = 11))


```
