---
title: "Time Series"
author: "Robert Dellinger"
date: "5/16/2022"
output: html_document
---

# Install Packages 

```{r setup, include=TRUE, message = FALSE}
knitr::opts_chunk$set(warning=FALSE, fig.algin="center")

library(lubridate)
library(tidyverse)
library(xtractomatic)
library(rerddap)
library(oce)
library(ocedata)
library(ncdf4)
library(lubridate)
library(tidync)
library(dplyr)
library(ggplot2)
library(raster)


```

# Visualizing Time Series Data

Data was downloaded from NOAA Environmental Research Division Data Access Program (EDRAPP) at a Southern California Coastal Ocean Observation Station (SCCOOS) linked here [https://sccoos.org/]. Data was downloaded from the station at the Santa Monica Pier. This data provides us with the SST measurements taken at frequent intervals for the nearshore coastal ocean of the Los Angeles region. 

Below we read in the packages using the nc_open() functionand extract time and SST variables from the data set

# Load in Data

```{r, Downloading Data (SCOSS) }

sccoss.data <- read_csv("/Users/robertdellinger/Documents/Research/Data/Temperature_Data/sccoss.santamonica.sst.2012.2022.csv")

names(sccoss.data) <- c("datetime","SST")

```

# Data Wrangling 

```{r, Data Wrangling SCOSS, warning= FALSE}

#clean the data
sccoss.data.clean  <- sccoss.data[-1,] %>% #remove first column
  mutate(datetime = str_remove(datetime, "Z"), #remove z from string
         date = as.Date(ymd(str_split_fixed(datetime, "T", n=Inf)[,1])), #split string into date
         time= str_split_fixed(datetime, "T", n=Inf)[,2], #split string into time
         SST = as.numeric(SST)) %>% #change sst to a numeric value
  filter(SST>10,
         SST<30) %>%
  dplyr::select("date", "SST") 

sccoss.summary <- sccoss.data.clean %>% 
  group_by(date)%>% # group date
  summarise(SST.mean = mean(SST, na.rm=TRUE), #calculate mean
            SST.sd = sd(SST, na.rm=TRUE),
            SST.min = min(SST, na.rm=TRUE),
            SST.max = max(SST, na.rm=TRUE))

sccoss.dataset <- full_join(sccoss.data.clean, sccoss.summary, by="date") #join data sets

sccoss.timeseries <- sccoss.dataset %>% 
  filter(date < ymd("2022-01-01"), #filter years
         date > ymd("2019-01-01"))

```

#Plotting Time Series Temperature Data

```{r, fig.cap="Sea Surface Temperature Time Series for Los Angeles"}

# make the plot and add labels
sst.plot <- ggplot(sccoss.timeseries, aes(x=date, y=SST.mean)) + #date by mean sst
  geom_line(aes(x=date, y=SST.mean), color="darkcyan", alpha=0.5) +
  geom_smooth(span=0.25, se=FALSE, color="darkcyan") +
  geom_ribbon(aes(ymin=SST.min, ymax=SST.max), fill="darkcyan", alpha=0.1)+
  ylab("Sea Surface Temperature (°C)") +  #add labels
  xlab("Date") +
labs(title = "Mean Sea Surface Temperature in Los Angeles, CA", subtitle = "Maximum, Minimum, & Mean SST Between 2019 & 2022 
(Data: Southern California Coastal Ocean Observation Station, Santa Monica Pier)") +
  theme_bw(base_size = 9) #

sst.plot

```