---
title: "Study Site Time Series"
author: "Robert Dellinger"
date: "5/16/2022"
output: html_document
---

# Install Packages 

```{r setup, include=TRUE, message = FALSE}
knitr::opts_chunk$set(warning=FALSE, fig.algin="center")

devtools::install_github("rmendels/xtractomatic")
library(lubridate)
library(tidyverse)
library(xtractomatic)
library(rerddap)
library(sp)
library(spData)
library(oce)
library(ocedata)
library(here)
```

Data was downloaded from https://shorestations.ucsd.edu/ for the shore station of Newport Beach, CA there are other stations for the span of the California Coast available to download on the website. Newport Beach Pier measurements collected by the City of Newport Beach Marine Operations Division.

Data Set: Newport Beach/Balboa Pier, 1925 – 2020.

# Sea Surface Temperature Heat Map Using Buoy Data off of Newport Beach, California. 

```{r, Downloading Data and Data Wrangling}

newport.temp.data <- read_csv(here("Data", "Site_Data", "scripps.newportbeach.sst.1925.2022.csv"))

names(newport.temp.data) <- c("year","month", "day", "NA1", "NA2", "SST", "NA3", "NA4")

newport.temp.data <- newport.temp.data[-c(1:29),] %>% 
  select(year, month, day, SST)

newport.data.clean <- newport.temp.data %>% 
  mutate(datetime = as.Date(paste(year, month, day, sep="-")),
         SST= as.numeric(SST)) %>% 
  filter(year >= 1925, year <= 2020) %>% 
  select(datetime, SST) %>% 
  arrange(datetime) %>% 
  na.omit()
  

```




#Interpolating Missing Values
```{r}

## define the total number of days in a year
days = 1:365
#define the total number of years as indicated by the data
years = 1925:2020

## create day and year variables from date
newport.data.clean = newport.data.clean %>% mutate(year = year(datetime), day_year = yday(datetime))

## interpolate the missing values
interpolated.temperature = interpBarnes(x = newport.data.clean$day_year, 
                                        y = newport.data.clean$year, 
                                        z = newport.data.clean$SST, 
                                        xgl = 365, 
                                        ygl = 96)

```

#
#Converting a Matrix to a Table and Time Series

```{r}

## extracted the matrix of interpolated values
interpolated.sst.mat = interpolated.temperature$zg

## tranform the matrix into the tabular form and combine the interpolated values with the raw
interpolated.sst.tb = interpolated.sst.mat %>% 
  as.tibble() %>% 
  gather(key = "year", value = "SST", 1:95)  %>%
  select(-year, sst.interp = SST) %>% 
  bind_cols()

# create a date sequence
dates = seq(as.Date("1925-01-01"), to=as.Date("2019-12-31"), "days")

#remove elap year dates
dates <- dates[!(format(dates,"%m") == "02" & format(dates, "%d") == "29"), drop = FALSE] #omit leap years


#Sequence the number of days throughout the time frame
interpolated.sst.tb = interpolated.sst.tb %>% 
  mutate(date=dates) %>% 
   select(date, sst.interp)


```



# Time Series 

The ts() function convert a numeric vector into an R time series object. The format is ts(vector, start=, end=, frequency=) where start and end are the times of the first and last observation and frequency is the number of observations per unit time (1=annual, 4=quartely, 12=monthly,365.24=daily, etc.). Our dataset contain daily sea surface temperature observations from 1997-01-01 to 2017-12-31. The daily time series object was created with the line code in the chunk below

```{r}


temperature.ts = ts(data = interpolated.sst.tb$sst.interp, 
                    start = c(1925,1), 
                    frequency = 365.24)

temperature.ts %>% plot(xlab = "", las = 1,
                  ylab = expression(Temperature~(degree*C)))

```


#Using ggplot 

```{r}

ggplot(data = interpolated.sst.tb, aes(x=date, y=sst.interp))+
  geom_line()+
  geom_smooth(span=0.25, color="red") +
  theme(axis.text = element_text(size = 11, colour = "black"),
        axis.title = element_text(size = 14, colour = "black"),
        panel.border = element_rect(colour = "black", fill = NA),
        panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(), 
        panel.grid.major.y = element_line(colour = "grey70", size = 0.25, linetype = 3)) +
  scale_y_continuous(breaks = seq(8, 24, 2)) +
  labs(x = "", y = expression(Temperature~(degree*C)))
```


# Salinity Buoy Data for Newport Beach, California. 

```{r}

newport.salinity.data <- read_csv(here("Data", "Site_Data", "scripps.newportbeach.salinity.1925.2022.csv"), skip=30, col_names = c("year","month", "day", "time", "NA", "salinity")) %>% 
  select(year, month, day, salinity)

newport.salinity.data.clean <- newport.salinity.data %>% 
  mutate(datetime = as.Date(paste(year, month, day, sep="-")),
         salinity= as.numeric(salinity)) %>% 
  filter(year >= 1925, year <= 2022,
         salinity >= 32, salinity <= 40) %>% 
  select(datetime, salinity) %>% 
  arrange(datetime) %>% 
  na.omit()
  

```


```{r}

ggplot(newport.salinity.data.clean, aes(x=datetime, y=salinity))+
  geom_line()+
  geom_smooth(span=0.25, color="lightblue") +
  theme(axis.text = element_text(size = 11, colour = "black"),
        axis.title = element_text(size = 14, colour = "black"),
        panel.border = element_rect(colour = "black", fill = NA),
        panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(), 
        panel.grid.major.y = element_line(colour = "grey70", size = 0.25, linetype = 3)) +
  scale_y_continuous(breaks = seq(30, 40, 1)) +
  labs(x = "Year", y = "Salinity (PSU)")

```


